Secure Composition \\ of Robust and Optimising Compilers

### 2023-09-28 csf/main.tex
crunch TODOs

\subsection{Applicability of the Compositionality Framework}

So far, this paper assumes a unified trace-model and secure compilers that do not change the behavior.
It is worth noting that in this setting, the theorem may not be applicable.
For example, consider $\rtp{\cc{\src{L}}{\trg{L}}}{\tmssafe}$ and $\rtp{\cc{\trg{L}}{\obj{L}}}{\smssafe}$, where $\cc{\trg{L}}{\obj{L}}$ introduces code that allocates a table that stores bounds information of pointers, but fails to insert code that deallocates this table.
By \Thmref{def:trace:tmsdef}, this violates $\tmssafe$.
Therefore, no behavior resulting from programs produced by the $\cc{\trg{L}}{\obj{L}}$ compiler is secure with respect to $\tmssafe$, but only for $\smssafe$.
$\cc{\trg{L}}{\obj{L}}$ modifies the behavior and because of this, the theorem is inapplicable.
\MPin{
	this is still applicable though
	many use it, and we showcase it too.
}

This section re-introduces secure compiler composition without this limitation. 
To this end, it utilises existing work on secure compilation~\cite{abate2021extacc} that defines Galois connections for several real-world compiler scenarios, such as a compilation pass that exploits undefined behavior.
This section is concerned about source-level properties, and henceforth 
\MPin{
	different source and target trace models
	.
	traces are related
	.
	there is a function tau that lefts the relation from individual traces to props/hprops
}
\MPin{
	drop next 3 defs
}

uses the existential image of a source-level property $\src{\pi}$ to map $\src{\pi}$ to a target-level property:

\begin{definition}[Existential Image of Properties]\label{def:eximg:prop}
  Given a cross-language trace relation $\sim$,
  $$
  \tau_\sim(\src{\pi}) \isdef \left\{
    \trg{\trace} \mid \exists\src{\trace}.\src{\trace}\sim\trg{\trace} \text{ and }
    \src{\trace}\in\src{\pi}
  \right\}
  $$
\end{definition}

\Thmref{def:eximg:prop} can be lifted to hyperproperties/classes as follows:
\begin{definition}[Existential Image of Hyperproperties]\label{def:eximg:hprop}
  Given a cross-language trace relation $\sim$,
  $$
  \tilde{\tau}_\sim(\src{\Pi}) \isdef \left\{
    \trg{\pi} \mid \exists\src{\pi}.\src{\pi}\in\src{\Pi} \text{ and }
    \trg{\pi}=\tau_\sim\left(\src{\pi}\right)
  \right\}
  $$
\end{definition}

\bul{A cross-language relation $\sim$ is well-formed with respect to $\src{\Pi}$} iff \rul{the existential images preserve membership}.
\begin{definition}[Well-formedness of Cross-Language Trace Relations for a Concrete Property $\src{\Pi}$]\label{def:wf:tracerel}
  $$
  \text{\bul{$\wf{\sim}{\src{\Pi}}$}} \isdef \text{\rul{$\forall \src{\pi}\in\src{\Pi}, \tau_\sim(\src{\pi})\in\tilde{\tau}_\sim(\src{\Pi})$}}
  $$
\end{definition}
\begin{definition}[Well-formedness of Cross-Language Trace Relations for a Class $\src{\class}$]\label{def:wfc:tracerel}
  $$
  \wfc{\sim}{\src{\class}} \isdef \forall \src{\Pi}\subseteq\src{\class}, \tau_\sim(\src{\Pi})\subseteq\tilde{\tau}_\sim(\src{\class})
  $$
\end{definition}
\MPin{
	explain with text and underline
}
\begin{lemma}[Well-Formedness Equivalence for Classes]\label{lem:wf:equiv:wfc}
  $\;\;$
  $ \wf{\sim}{\src{\class}} \Leftrightarrow\ \wfc{\sim}{\src{\class}} $\Coqed
\end{lemma}

\Thmref{def:rtp} needs to be adapted to account for different target-level trace-models:

@@

To state a similar theorem as \Thmref{thm:rtp} but for \Cref{def:rtpsim}, it is necessary to define composition of cross-language trace relations:
\MPin{
	do in text
}
\begin{definition}[Composing Cross-Language Trace Relations]
  Given $\sim_1 : \src{\trace}\to\irl{\trace}\to\text{Prop},\sim_2 : \irl{\trace}\to\trg{\trace}\to\text{Prop}$, then 
  $$  
  \sim_1\bullet\sim_2\ \isdef \lambda \src{\trace}\ \trg{\trace}. \exists \irl{\trace}. \src{\trace}\sim_1\irl{\trace} \text{ and } \irl{\trace}\sim_2\trg{\trace}
  $$
\end{definition}

\MPin{
	explain
}
\begin{theorem}[Sequential Composition of Secure Compilers]\label{thm:rtpsim}
  If $\rtpsim{\cc{\src{L}}{\trg{L}}}{\class[_{1}]}{\sim_1}$, $\rtpsim{\cc{\trg{L}}{\obj{L}}}{\tilde{\tau}_{\sim_1}\left(\class[_{2}]\right)}{\sim_2}$, and $\wf{\sim_1}{\class[_2]}$, then $\rtpsim{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\class[_{1}]\cap\class[_{2}]}{\sim_1\bullet\sim_2}$. \Coqed
\end{theorem}

Reconsidering the example from the beginning of the section, suppose $\sim_{{\smssafe}}$ is a suitable trace relation such that $\trg{\trace}\sim_{{\smssafe}}\irl{\trace}$ establishes the fact that $\irl{\trace}$ contains an allocation of a table containing bounds information which is never deallocated.
It is impossible that $\wf{\sim_{{\smssafe}}}{{\tmssafe}}$, since every behavior of compiled $\trg{L}$ programs never contains a deallocation of the bounds-information table.
Still, supposing $\rtpsim{\cc{\src{L}}{\trg{L}}}{\tmssafe}{{\sim_{\tmssafe}}}$, $\rtpsim{\cc{\trg{L}}{\irl{L}}}{\smssafe}{{\sim_{\smssafe}}}$,
and $\wf{{\sim_{\tmssafe}}}{{\smssafe}}$ allows to apply \Thmref{thm:rtpsim}.
However, the resulting class $\tilde{\tau}_{{{\sim_{\tmssafe}}\bullet{\sim_{\smssafe}}}}\left(\tmssafe\cap\smssafe\right)=\emptyset$, which the following lemma shows to hold in general.

@@

\MPin{
	drop
}
With this, the previous claim follows as a simple corollary:
\begin{corollary}
  $\tilde{\tau}_{{{\sim_{\tmssafe}}\bullet{\sim_{\smssafe}}}}\left(\tmssafe\cap\smssafe\right)=\emptyset$.
\end{corollary}
\begin{proof}
  Note that $\neg\left(\wf{\sim_{{\smssafe}}}{{\tmssafe}}\right)$, since the bounds-table is never deallocated and therefore, for any $\trg{\trace}\in\tmssafe$, $\tau_{\sim_{\tmssafe}}\left(\trg{\trace}\right)\notin\tilde{\tau}_{\sim_{\tmssafe}}\left({\tmssafe}\right)$.
  However, by assumption, $\wf{\sim_{{\tmssafe}}}{{\tmssafe}}$.
  Using \Thmref{lem:useless} it follows that $\tilde{\tau}_{{{\sim_{\tmssafe}}\bullet{\sim_{\smssafe}}}}\left(\tmssafe\right)=\emptyset$.
\end{proof}

\MPin{
	what about sim2 and C1?
	-> no relation 
}
\MPin{
	For the sake of simplicity we go back to the singel trace model and go back to the framework of single trace model
}

@@

\vspace{-1em}
\begin{gather*}\small
  \begin{aligned}
    \cca(\src{x}) = &\ \trg{x} 
    % \\
  	&
    \cca(\src{\lbinop{\expr[_{1}]}{\expr[_{2}]}}) = &\ \lbinop[\trg]{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr[_{2}]})\right]} \\
    \cca(\src{n}) = &\ \trg{n} 
    % \\
    &
    \cca(\src{\lget{x}{\expr}}) = &\ \lget[\trg]{\trg{x}}{\left[\cca(\src{\expr})\right]}
  \end{aligned}
    \\
  \begin{aligned}
    \cca(\src{\ldelete{x}}) = &\ \ldelete[\trg]{\left[\cca(\src{x})\right]} \\
%    \cca(\src{\lset{x}{\expr[_{1}]}{\expr[_{2}]}}) & = \lset[\trg]{\left[\cca(\src{x})\right]}{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr_{2}})\right]} \\
\cca(\src{\lfunction{g}{x:\natt\to\type[_{e}]}{\expr}})  = &\ \lfunction[\trg]{\trg{g}}{\trg{x}}{\lifz[\trg]{\trg{\lhast{x}{\natt}}}{\\%
                                                            &\left[\cca(\src{\expr})\right] %
                                                                                                 }{\labort[\trg]}}
  \end{aligned}
\end{gather*}

@@

>>>>

\subsection{Applicability of the Compositionality Framework}\label{subsec:app}

So far, this paper assumes a unified trace-model and secure compilers that do not change the behavior.
It is worth noting that, as demonstrated in our case study, quite a few compilation passes do not change the behavior in such a way that the projections to the specification trace model become unequal to each other, i.e., source and target language of the compiler essentially have a different trace model. 
However, there are instances of passes where this is not the case.
For example, consider $\rtp{\cc{\src{L}}{\trg{L}}}{\tmssafe}$ and $\rtp{\cc{\trg{L}}{\obj{L}}}{\smssafe}$, where $\cc{\trg{L}}{\obj{L}}$ introduces code that allocates a table that stores bounds information of pointers, but fails to insert code that deallocates this table.
By \Thmref{def:trace:tmsdef}, this violates $\tmssafe$.
Therefore, no behavior resulting from programs produced by the $\cc{\trg{L}}{\obj{L}}$ compiler is secure with respect to $\tmssafe$, but only for $\smssafe$.
$\cc{\trg{L}}{\obj{L}}$ modifies the behavior and this section re-introduces the compositionality framework with support for these situations.

To this end, the section utilises existing work on secure compilation~\cite{abate2021extacc} that defines Galois connections for several real-world compiler scenarios, such as a compilation pass that exploits undefined behavior.
This section is concerned about a compilation setting with different source and target trace models. 
To this end, there needs to be a way to say that two traces $\src{\trace}$ and $\trg{\trace}$ are related ($\sim$) and a way to translate a source-level property $\src{\pi}$ into a target-level property ($\tau_\sim\left( \src{\pi} \right)$~\cite{abate2021extacc}) using this trace relation.
Moreover, the function to map source-level properties to target-level properties can be lifted to translate source-level classes of properties $\src{\class}$ to target-level classes ($\tilde{\tau}_\sim\left(\src{\class}\right)$).
The section refers to these functions as existential images.

With this, \bul{a cross-language relation $\sim$ is well-formed with respect to a class of source-level properties $\src{\class}$} iff \rul{the existential images preserve set inclusion}.
\begin{definition}[Well-formedness of Cross-Language Trace Relations for a Class $\src{\class}$]\label{def:wfc:tracerel}
  $$
  \text{\bul{$\wfc{\sim}{\src{\class}}$}} \isdef \text{\rul{$\forall \src{\Pi}\subseteq\src{\class}, \tau_\sim(\src{\Pi})\subseteq\tilde{\tau}_\sim(\src{\class})$}}
  $$
\end{definition}

\Thmref{def:rtp} needs to be adapted to account for different target-level trace-models~\cite{abate2021extacc}:

@@

To state a similar theorem as \Thmref{thm:rtp} but for \Cref{def:rtpsim}, it is necessary to define composition of cross-language trace relations.
Suppose there are traces $\src{\trace}$ and $\trg{\trace}$ as well as two trace relations $\sim_1$ and $\sim_2$. 
The traces are related $\src{\trace}\sim_1\bullet\sim_2\trg{\trace}$ if there is another trace $\irl{\trace}$ such that $\src{\trace}\sim_1\irl{\trace}$ and $\irl{\trace}\sim_2\trg{\trace}$.

For the adapted sequential composition theorem, let \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserve the class $\src{\class[_1]}$ under $\sim_1$} and let \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserve the class $\tilde{\tau}_{\sim_1}\left(\src{\class[_2]}\right)$ under $\sim_2$}.
Then, \iul{when the cross-language relation $\sim_1$ is well-formed with respect to class $\src{\class[_2]}$}, it follows that \oul{the composed compiler $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\src{\class[_1]}\cap\src{\class[_2]}$ under $\sim_1\bullet\sim_2$}.
\begin{theorem}[Sequential Composition of Secure Compilers]\label{thm:rtpsim}
  If \bul{$\rtpsim{\cc{\src{L}}{\trg{L}}}{\class[_{1}]}{\sim_1}$}, \rul{$\rtpsim{\cc{\trg{L}}{\obj{L}}}{\tilde{\tau}_{\sim_1}\left(\class[_{2}]\right)}{\sim_2}$}, and \iul{$\wf{\sim_1}{\class[_2]}$}, then \oul{$\rtpsim{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\class[_{1}]\cap\class[_{2}]}{\sim_1\bullet\sim_2}$}. \Coqed
\end{theorem}

For space reasons, \Thmref{thm:urtp} and \Thmref{thm:lrtp} are not presented here, but can be found in the technical appendix ($\CoqSymbol$).
Note that it is not necessary to require $\sim_2$ to be well-formed with respect to $\class[_1]$.

However, reconsidering the example from the beginning of the section, suppose $\sim_{{\smssafe}}$ is a suitable trace relation such that $\trg{\trace}\sim_{{\smssafe}}\irl{\trace}$ establishes the fact that $\irl{\trace}$ contains an allocation of a table containing bounds information which is never deallocated.
It is impossible that $\wf{\sim_{{\smssafe}}}{{\tmssafe}}$, since every behavior of compiled $\trg{L}$ programs never contains a deallocation of the bounds-information table.
Still, supposing $\rtpsim{\cc{\src{L}}{\trg{L}}}{\tmssafe}{{\sim_{\tmssafe}}}$, $\rtpsim{\cc{\trg{L}}{\irl{L}}}{\smssafe}{{\sim_{\smssafe}}}$,
and $\wf{{\sim_{\tmssafe}}}{{\smssafe}}$ allows to apply \Thmref{thm:rtpsim}.
However, the resulting class $\tilde{\tau}_{{{\sim_{\tmssafe}}\bullet{\sim_{\smssafe}}}}\left(\tmssafe\cap\smssafe\right)=\emptyset$, which the following lemma shows to hold in general.

@@

For sake of simplicity, the case-study presented in the following sections does not consider different trace-models: The traces of all programs of the presented languages are projected to a unified trace model. 
The case-study demonstrates that, while the compositionality framework without the consideration for different trace-models may seem overly simplistic, it is useful and simpler in practice, if applicable.

@@

\begin{gather*}\small
  \begin{aligned}
    \cca(\src{x}) = &\ \trg{x} 
    % \\
  	&
    \cca(\src{\lbinop{\expr[_{1}]}{\expr[_{2}]}}) = &\ \lbinop[\trg]{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr[_{2}]})\right]} \\
    \cca(\src{n}) = &\ \trg{n} 
    % \\
    &
    \cca(\src{\lget{x}{\expr}}) = &\ \lget[\trg]{\trg{x}}{\left[\cca(\src{\expr})\right]}
  \end{aligned}
    \\
  \begin{aligned}
    \cca(\src{\ldelete{x}}) = &\ \ldelete[\trg]{\left[\cca(\src{x})\right]} \\
%    \cca(\src{\lset{x}{\expr[_{1}]}{\expr[_{2}]}}) & = \lset[\trg]{\left[\cca(\src{x})\right]}{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr_{2}})\right]} \\
\cca(\src{\lfunction{g}{x:\natt\to\type[_{e}]}{\expr}})  = &\ \lfunction[\trg]{\trg{g}}{\trg{x}}{\lifz[\trg]{\trg{\lhast{x}{\natt}}}{\\%
                                                            &\left[\cca(\src{\expr})\right] %
                                                                                                 }{\labort[\trg]}}
  \end{aligned}
\end{gather*}

@@

\section{Auxiliary definitions for \Cref{subsec:app}}
\begin{definition}[Existential Image of Properties]\label{def:eximg:prop}
  Given a cross-language trace relation $\sim$,
  $$
  \tau_\sim(\src{\pi}) \isdef \left\{
    \trg{\trace} \mid \exists\src{\trace}.\src{\trace}\sim\trg{\trace} \text{ and }
    \src{\trace}\in\src{\pi}
  \right\}
  $$
\end{definition}

\Thmref{def:eximg:prop} can be lifted to hyperproperties/classes as follows:
\begin{definition}[Existential Image of Hyperproperties]\label{def:eximg:hprop}
  Given a cross-language trace relation $\sim$,
  $$
  \tilde{\tau}_\sim(\src{\Pi}) \isdef \left\{
    \trg{\pi} \mid \exists\src{\pi}.\src{\pi}\in\src{\Pi} \text{ and }
    \trg{\pi}=\tau_\sim\left(\src{\pi}\right)
  \right\}
  $$
\end{definition}
\bul{A cross-language relation $\sim$ is well-formed with respect to a source-level (hyper-)property $\src{\Pi}$} iff \rul{the existential images preserve membership}.
\begin{definition}[Well-formedness of Cross-Language Trace Relations for a Properties $\src{\Pi}$]\label{def:wf:tracerel}
  $$
  \text{\bul{$\wfc{\sim}{\src{\Pi}}$}} \isdef \text{\rul{$\forall \src{\pi}\in\src{\Pi}, \tau_\sim(\src{\pi})\in\tilde{\tau}_\sim(\src{\Pi})$}}
  $$
\end{definition}
\begin{lemma}[Well-Formedness Equivalence for Classes]\label{lem:wf:equiv:wfc}
  $\;\;$
  $ \wf{\sim}{\src{\class}} \Leftrightarrow\ \wfc{\sim}{\src{\class}} $\Coqed
\end{lemma}
